No one-size-fits-all solution to clean GBIF

### 2020-08-04 manuscript_peerj.Rmd
2nd resubmission

  Species occurrence records provide the basis for many biodiversity studies. They derive from georeferenced specimens deposited in natural history collections and visual observations, such as those obtained through various mobile applications. Given the rapid increase in availability of such data, the control of quality and accuracy constitutes a particular concern. Automatic filtering is a scalable and reproducible means to identify potentially problematic records and tailor datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org), for biodiversity analyses. However, it is unclear how much data may be lost by filtering, whether the same filters should be applied across all taxonomic groups, and what the effect of filtering is on common downstream analyses. Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 18 Neotropical taxa, including terrestrial and marine animals, fungi, and plants downloaded from GBIF. We find that a total of 44.3% of the records are potentially problematic, with large variation across taxonomic groups (25 - 90%). A small fraction of records was identified as erroneous in the strict sense (4.2%), and a much larger proportion as unfit for most downstream analyses (41.7%). Filters of duplicated information, collection year, and basis of record, as well as coordinates in urban areas, or for terrestrial taxa in the sea or marine taxa on land, have the greatest effect. Automated filtering can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus. Our results stress the importance of thorough recording and exploration of the meta-data associated with species records for biodiversity research.

@@

@@

<!-- # Keywords -->
<!-- Automated cleaning, automated conservation assessment, data quality, GBIF, Neotropics. -->

@@

# Introduction {-}
Publicly available species distribution data have become a crucial resource in biodiversity research, including studies in ecology, biogeography, systematics and conservation biology. In particular, the availability of digitized collections from museums and herbaria and citizen science observations has increased drastically over the last few years. As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.4 billion geo-referenced occurrence records for species from across the globe and the tree of life.

A central challenge to the use of these publicly available species occurrence data in research is problematic geographic coordinates, which are either erroneous or unfit for downstream analyses [for instance because they are overly imprecise, @Anderson2016]. Problems mostly arise because data aggregators such as GBIF integrate records collected with different methodologies in different places at different times--- often without centralized curation and only rudimentary meta-data. For instance, problematic coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common [@Maldonado2015; @Yesson2007] and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in cultivation or captivity [@Zizka2019].

@@

# Material and Methods {-}
## Choice of study taxa {-}
This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 which gathered students and researchers working with different taxonomic groups of animals, fungi, and plants across the Neotropics. Each participant analysed geographic occurrence data from their taxonomic group of interest and commented on the results for their group. Hence, we include groups based on the expertise of the participants rather than following an arbitrary choice of taxa and taxonomic ranks. We acknowledge a varying degree of documented expertise and number of years working on each group. We obtained public occurrence records for 18 taxa, including one plant family, nine plant genera, one genus of fungi, three families and one genus of terrestrial arthropods, one family of snakes, one family of skates, and one genus of bony fish (Table 1).

@@

We downloaded occurrence information for all study groups from www.gbif.org using the `rgbif` v1.4.0 package [@Chamberlain2017] in R (GBIF.org, 2019a-p,2020a,b). We downloaded GBIF-interpreted data including only records with geographic coordinates and limited the study area to a rectangle between 90$^\circ$S - 33$^\circ$ N and 35$^\circ$ W - 120$^\circ$ W reflecting the Neotropics [@Morrone2014], our main area of expertise. The natural distributions of all included taxa are confined to the Neotropics except for Arhynchobatidae, Diogenidae, Dipsadidae, Entomobryidae, *Gaylussacia*, Iridaceae, Neanuridae, and *Tillandsia*, for which we only obtained the Neotropical occurrences. We consider GBIF data generally of high quality and use them as a case study because GBIF is the largest, most widely used and taxonomically most comprehensive data source for species occurrence records; however many more exist [e.g., https://bien.nceas.ucsb.edu/bien/, www.fishbase.de or @Guedes2018]. The variety of existing databases is also the reason why we downloaded data without accounting for GBIF-internal quality testing ("Geospatial issues"), since quality control mechanisms might differ among databases and sometimes quality control might not be carried out (for instance in *de novo* databases compiled by groups of researchers).

##  Automated cleaning {-}
We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the raw data downloaded from GBIF ("raw", hereafter) using meta-data for those records for which they were available [although meta-data were often missing, @Peterson2018], removing: 1) records with a coordinate precision below 100 km (as this represents the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99. Furthermore, we rounded the geographic coordinates to four decimal places and retained only one record per species per location (i.e., test for duplicated records). In a second step, we used the `clean_coordinates` function of the `CoordinateCleaner v2.0-11` package [@Zizka2019] with default options to flag errors that are common to biological data sets (“filtered”, hereafter). These include: coordinates in the sea for terrestrial taxa and on land for marine taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions. See Table 2 for a summary of all filters we used and their classification into "erroneous" and "unfit".

@@

We retrieved a total of `r nrow(dat)` species occurrence records, with a median of `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist() %>%  round(0)` records per study group and `r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per species (Table 3, Appendix 1). We obtained most records for `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(taxon) %>%  unlist()` (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs) %>%  unlist()`) and fewest for *`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(taxon) %>%  unlist()`* (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(nrecs) %>%  unlist()`). The species with most records was *`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(species)`* (`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs)`).

Our automated tests filtered a total of `r sum(!dat$summary)` records (Fig. \ref{fig:total}, erroneous: `r err`, unfit: `r unfit`), with a median of `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per group (erroneous: `r median(tab2$total_error)`%, unfit: `r median(tab2$total_unfit)`%). Overall, the most important test was for duplicated records (on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!.dpl)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per taxonomic group). The filtering steps based on record meta-data that filtered the largest number of records were the basis of records (`r dat %>% group_by(taxon) %>% summarize(num = mean(!coordinate_base)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the collection year (`r dat %>% group_by(taxon) %>% summarize(num = mean(!record_age)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%). The most important automated tests were for urban area (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.urb)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the occurrence from records of terrestrial taxa in the sea and marine taxa on land (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.sea)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%, see Table 3 and Appendix 1 in the electronic supplement for further details and the absolute numbers).  Only a few records were filtered by the coordinate precision, zero coordinates and biodiversity institution tests (Fig. \ref{fig:split}).

`r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(1) %>% unlist()`, `r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(2) %>% unlist()`, and `r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(3) %>% unlist()` had the highest fraction of filtered records (Table 3). In general, the different filters we tested were of similar importance for different study groups. There were few outstanding exceptions, including the particularly high proportions of records filtered by the "basis of record test" for *Tityus* (7.0%), Dipsadidae (5.6%), *Prosthechea* (5.0%) and *Tillandsia* (4.9%), by the collection year for Dipsadidae (11.3%), by the taxonomic identification level for  *Tityus* (1.6%), by the capital coordinates for *Oocephalus* (6.1%) and *Gaylussacia* (3.2%), by the seas/land test for Diogenidae and *Thozetella*, and by the urban areas test for *Oocephalus* (13.3%) and Iridaceae (12.3%). Furthermore, Entomobryidae differed considerably from all other study taxa with exceptionally high numbers of records filtered by the "basis of record", "level of identification" and "urban areas" tests.

@@

We found IUCN assessments for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>% nrow()` species that were also included in our distribution data from 11 of our study groups (Table 4, Appendix 3). The fraction of species evaluated varied among the study group, with a maximum of `r tab3 %>% select(iucn_frac_eval) %>% unlist() %>%  max()`% for *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>% slice(1L) %>% select(taxon) %>%  unlist()`* and *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>%  select(taxon) %>% slice(2) %>% unlist()`* and a minimum of `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(iucn_frac_eval) %>%  unlist()`% for `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(taxon) %>%  unlist()` (note that the number of total species varied considerably among groups). The median percentage of species per study group with an IUCN assessment was `r tab3 %>% select(iucn_frac_eval) %>% unlist() %>%  median()`%. A total of `r aa %>% filter(!is.na(iucn_category)) %>%  filter(iucn_category != "DD") %>%filter(iucn_category != "EX") %>%  mutate(test = iucn_status == "possibly_threatened") %>% select(test) %>% unlist() %>%  sum()` species were listed as *Threatened* by the IUCN global Red List (CR = `r aa %>% filter(iucn_category == "CR") %>% nrow()`, EN = `r aa %>% filter(iucn_category == "EN") %>% nrow()`, VU = `r aa %>% filter(iucn_category == "VU") %>% nrow()`) and `r  aa %>% filter(!is.na(iucn_category)) %>%  filter(iucn_category != "DD") %>%filter(iucn_category != "EX") %>%  mutate(test = iucn_status == "not_threatened") %>% select(test) %>% unlist() %>%  sum()` as *Not Threatened*.

We obtained automated conservation assessments for `r aa %>% select(flag_Category_CriteriaB) %>% filter(!is.na(flag_Category_CriteriaB)) %>% nrow()` species in the filtered dataset. Based on the filtered data, the automated conservation assessment evaluated `r sum(aa$flag_status == "possibly_threatened", na.rm = TRUE)` species as possibly threatened (`r round(sum(aa$flag_status == "possibly_threatened", na.rm=TRUE) / aa %>% filter(!is.na(flag_status)) %>% nrow() * 100, 1)`%, CR = `r aa %>% filter(flag_Category_CriteriaB == "CR") %>% nrow()`, EN = `r aa %>% filter(flag_Category_CriteriaB == "EN") %>% nrow()`, VU = `r aa %>% filter(flag_Category_CriteriaB == "VU") %>% nrow()`, see Appendix 3 for assessments of all species). The automated assessment based on the filtered dataset agreed with the IUCN assessment for identifying species as possibly threatened (CR, EN, VU) for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status)) %>% mutate(agreement = iucn_status == flag_status) %>% select(agreement) %>% unlist() %>% sum() ` species (`r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status))  %>% mutate(agreement = iucn_status == flag_status) %>% select(agreement) %>% unlist() %>% mean() %>%  round(3) * 100`%; Table 4). Filtering reduced the EOO by `r tab3 %>%  select(flag_median_EOO_change) %>%  unlist() %>%  median() %>%  abs()`% and the AOO by `r tab3 %>%  select(flag_median_AOO_change) %>%  unlist() %>%  median() %>% abs()`% on median per group. For the raw dataset the agreement with IUCN was higher at `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% mutate(agreement = iucn_status == raw_status) %>% select(agreement) %>% unlist() %>% sum() ` species (`r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% mutate(agreement = iucn_status == raw_status) %>% select(agreement) %>% unlist() %>% mean() %>%  round(3) * 100`%).

# Discussion {-}
<!-- add some subheadings to discussion -->
Automated flagging based on meta-data and automatic tests filtered on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% of the records per taxonomic group; `r min(tab2$total_unfit)`%-`r max(tab2$total_unfit)`% as "unfit" and `r min(tab2$total_error)`%-`r max(tab2$total_error)`% as "erroneous". The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (**Question 1**). The importance of different tests was similar across taxonomic groups, with particular exceptions for the tests on basis of record, collection year, capital coordinates, and urban areas (**Question 2**). The results for species richness were similar between the raw and filtered data with some improvements using the filters. We found little impact of filtering on the accuracy of the automated conservation assessments (**Question 3**).

@@

Overall, we consider the effect of the automated filters as positive since they identified the above-mentioned issues and increased the data precision and reduced computational burden (Table 3, Appendix 2). However, in some cases filters failed to remove major issues, often due to incomplete metadata. For instance, for Diogenidae we found at least two records of an extinct species (*Paguristes mexicanus*) which slipped the "basis of record" test because they were marked as "preserved specimen" rather than "fossil specimen". Furthermore, for Entomobryidae we found that for `r dat %>% filter(taxon == "Entomobryidae") %>%  filter(!is.na(species)) %>%  filter(taxonRank == "UNRANKED") %>%  nrow()` records the meta-data on taxonomic rank was "UNRANKED" despite all of them being identified to species level, leading to a high fraction of records removed by the "Identification level" filter. Additionally automated filters might be overly strict  or unsuitable for certain taxa. For instance, in Entomobryidae, `r dat %>% filter(taxon == "Entomobryidae") %>% filter( basisOfRecord == "MATERIAL_SAMPLE") %>% nrow()` samples were marked as material samples and therefore removed by our global filter retaining only specimen and observation data, which in this case was overly strict.

The filters we included in this study address a set of important but relatively easy to identify problems. In fact, the internal quality control of GBIF does flag some of the problems we tested for (i.e., zero coordinates, equal lat/lon) while others might be implemented in the near future (country centroids, https://data-blog.gbif.org/post/country-centroids/, https://data-blog.gbif.org/post/country-centroids/). While this internal quality is very helpful, we see a huge potential to overcome issues with data quality in a user-feedback system that allows users to provide expert assessments, i.e. a meta-annotation of records being challenged (and why). Such a system would not need to change the original data and could include multiple levels to account for differing opinions.

As a next step for automated filtering, it is possible to detect additional problematic records with the help of external species distribution information, which is available for some taxa, for example from www.iucn.org (vertebrates) or https://wcsp.science.kew.org/ (selected seed plant families). Furthermore, while in this study we focused on meta-data and geographic filtering, taxonomic cleaning---the resolution of synonymies and identification of accepted names---is another important part of data curation, but depends on taxon-specific taxonomic backbones and synonymy lists which are not readily available for many groups and often are contradictory within individual taxa.

@@

# Conclusions {-}
Our results suggest that between one to half of the occurrence records obtained from GBIF might be unsuitable for downstream biodiversity analyses. While the majority of these records might not be erroneous *per se*, they are overly imprecise and thereby increase uncertainty of downstream results or add computational burden on big data analyses.

>>>>

  Species occurrence records provide the basis for many biodiversity studies. They derive from georeferenced specimens deposited in natural history collections and visual observations, such as those obtained through various mobile applications. Given the rapid increase in availability of such data, the control of quality and accuracy constitutes a particular concern. Automatic filtering is a scalable and reproducible means to identify potentially problematic records and tailor datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org), for biodiversity analyses. However, it is unclear how much data may be lost by filtering, whether the same filters should be applied across all taxonomic groups, and what the effect of filtering is on common downstream analyses. Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 18 Neotropical taxa, including terrestrial and marine animals, fungi, and plants downloaded from GBIF. We find that a total of 44.3% of the records are potentially problematic, with large variation across taxonomic groups (25 - 90%). A small fraction of records are identified as erroneous in the strict sense (4.2%), and a much larger proportion as unfit for most downstream analyses (41.7%). Filters of duplicated information, collection year, and basis of record, as well as coordinates in urban areas, or for terrestrial taxa in the sea or marine taxa on land, have the greatest effect. Automated filtering can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus. Our results stress the importance of thorough recording and exploration of the meta-data associated with species records for biodiversity research.

@@

  - \usepackage{float}
  - \floatplacement{figure}{H} 

@@

@@

# Introduction {-}
Publicly available species distribution data have become a crucial resource in biodiversity research, including studies in ecology, biogeography, systematics and conservation biology. In particular, the availability of digitized collections from museums and herbaria and citizen science observations has increased drastically over the last few years. As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.5 billion geo-referenced occurrence records for species from across the globe and the tree of life.

A central challenge to the use of these publicly available species occurrence data in research is problematic geographic coordinates, which are either erroneous or unfit for downstream analyses [for instance because they are overly imprecise, @Anderson2016]. Problems mostly arise because data aggregators such as GBIF integrate records collected with different methodologies in different places at different times---often without centralized curation and only rudimentary meta-data. For instance, problematic coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common [@Maldonado2015; @Yesson2007] and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in cultivation or captivity [@Zizka2019].

@@

# Material and Methods {-}
## Choice of study taxa {-}
This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 which gathered students and researchers working with different taxonomic groups of animals, fungi, and plants across the Neotropics (Fig. \ref{fig:species}). Each participant analysed geographic occurrence data from their taxonomic group of interest and commented on the results for their group. Hence, we include groups based on the expertise of the participants rather than following an arbitrary choice of taxa and taxonomic ranks. We acknowledge a varying degree of documented expertise and number of years working on each group. We obtained public occurrence records for 18 taxa, including one plant family, nine plant genera, one genus of fungi, three families and one genus of terrestrial arthropods, one family of snakes, one family of skates, and one genus of bony fish (Table 1).

@@

We downloaded occurrence information for all study groups from www.gbif.org using the `rgbif` v1.4.0 package [@Chamberlain2017] in R (GBIF.org, 2019a-p,2020a,b). We downloaded GBIF-interpreted data including only records with geographic coordinates and limited the study area to a rectangle between 90$^\circ$S - 33$^\circ$ N and 35$^\circ$ W - 120$^\circ$ W reflecting the Neotropics [@Morrone2014], our main area of expertise. The natural distributions of all included taxa are confined to the Neotropics except for Arhynchobatidae, Diogenidae, Dipsadidae, Entomobryidae, *Gaylussacia*, Iridaceae, Neanuridae, and *Tillandsia*, for which we only obtained the Neotropical occurrences. We consider GBIF data generally of high quality and use them as a case study because GBIF is the largest, most widely used and taxonomically most comprehensive data source for species occurrence records; however many more exist [e.g., https://bien.nceas.ucsb.edu/bien/, www.fishbase.de or @Guedes2018]. GBIF provides information on the internal consistency of records, among others including information on decimal rounding of coordinates, geographic projection and date validity and geospatial issues. Since we specifically aimed to test the effect of user-level filtering we included records flagged with issues by GBIF (this was also the default option). Geospatial issues flagged by GBIF only concerned 0.4% of the records used in this study and including them had the added benefit to make our results directly comparable to other databases, which may use different internal consistency checks or none at all.

##  Automated cleaning {-}
We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the data as downloaded from GBIF ("raw", hereafter) using meta-data for those records for which they were available [although meta-data were often missing, @Peterson2018], removing: 1) records with a coordinate precision below 100 km (as this represents the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99. Furthermore, we rounded the geographic coordinates to four decimal places and retained only one record per species per location (i.e., test for duplicated records). In a second step, we used the `clean_coordinates` function of the `CoordinateCleaner v2.0-11` package [@Zizka2019] with default options to flag errors that are common to biological data sets (“filtered”, hereafter). These include: coordinates in the sea for terrestrial taxa and on land for marine taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions. See Table 2 for a summary of all filters we used and their classification into "erroneous" and "unfit".

@@

We retrieved a total of `r nrow(dat)` species occurrence records, with a median of `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist() %>%  round(0)` records per study group and `r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per species (Table 3, Appendix 1). We obtained most records for `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% dplyr::select(taxon) %>%  unlist()` (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% dplyr::select(nrecs) %>%  unlist()`) and fewest for *`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% dplyr::select(taxon) %>%  unlist()`* (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% dplyr::select(nrecs) %>%  unlist()`). The species with most records was *`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% dplyr::select(species)`* (`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% dplyr::select(nrecs)`).

Our automated tests filtered a total of `r sum(!dat$summary)` records (Fig. \ref{fig:total}, erroneous: `r err`, unfit: `r unfit`), with a median of `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  dplyr::select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per group (erroneous: `r median(tab2$total_error)`%, unfit: `r median(tab2$total_unfit)`%). Overall, the most important test was for duplicated records (on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!.dpl)) %>%  dplyr::select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per taxonomic group). The filtering steps based on record meta-data that filtered the largest number of records were the basis of records (`r dat %>% group_by(taxon) %>% summarize(num = mean(!coordinate_base)) %>%  dplyr::select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the collection year (`r dat %>% group_by(taxon) %>% summarize(num = mean(!record_age)) %>%  dplyr::select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%). The most important automated tests were for urban area (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.urb)) %>%  dplyr::select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the occurrence from records of terrestrial taxa in the sea and marine taxa on land (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.sea)) %>%  dplyr::select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%, see Table 3 and Appendix 1 in the electronic supplement for further details and the absolute numbers).  Only a few records were filtered by the coordinate precision, zero coordinates and biodiversity institution tests (Fig. \ref{fig:split}).

`r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% dplyr::select(taxon) %>% slice(1) %>% unlist()`, `r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% dplyr::select(taxon) %>% slice(2) %>% unlist()`, and `r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% dplyr::select(taxon) %>% slice(3) %>% unlist()` had the highest fraction of filtered records (Table 3). In general, the different filters we tested were of similar importance for different study groups. There were few outstanding exceptions, including the particularly high proportions of records filtered by the "basis of record test" for *Tityus* (7.0%), Dipsadidae (5.6%), *Prosthechea* (5.0%) and *Tillandsia* (4.9%), by the collection year for Dipsadidae (11.3%), by the taxonomic identification level for  *Tityus* (1.6%), by the capital coordinates for *Oocephalus* (6.1%) and *Gaylussacia* (3.2%), by the seas/land test for Diogenidae and *Thozetella*, and by the urban areas test for *Oocephalus* (13.3%) and Iridaceae (12.3%). Furthermore, Entomobryidae differed considerably from all other study taxa with exceptionally high numbers of records filtered by the "basis of record", "level of identification" and "urban areas" tests.

@@

We found IUCN assessments for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>% nrow()` species that were also included in our distribution data from 11 of our study groups (Table 4, Appendix 3). The fraction of species evaluated varied among the study group, with a maximum of `r tab3 %>% dplyr::select(iucn_frac_eval) %>% unlist() %>%  max()`% for *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>% slice(1L) %>% dplyr::select(taxon) %>%  unlist()`* and *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>%  dplyr::select(taxon) %>% slice(2) %>% unlist()`* and a minimum of `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  dplyr::select(iucn_frac_eval) %>%  unlist()`% for `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  dplyr::select(taxon) %>%  unlist()` (note that the number of total species varied considerably among groups). The median percentage of species per study group with an IUCN assessment was `r tab3 %>% dplyr::select(iucn_frac_eval) %>% unlist() %>%  median()`%. A total of `r aa %>% filter(!is.na(iucn_category)) %>%  filter(iucn_category != "DD") %>%filter(iucn_category != "EX") %>%  mutate(test = iucn_status == "possibly_threatened") %>% dplyr::select(test) %>% unlist() %>%  sum()` species were listed as *Threatened* by the IUCN global Red List (CR = `r aa %>% filter(iucn_category == "CR") %>% nrow()`, EN = `r aa %>% filter(iucn_category == "EN") %>% nrow()`, VU = `r aa %>% filter(iucn_category == "VU") %>% nrow()`) and `r  aa %>% filter(!is.na(iucn_category)) %>%  filter(iucn_category != "DD") %>%filter(iucn_category != "EX") %>%  mutate(test = iucn_status == "not_threatened") %>% dplyr::select(test) %>% unlist() %>%  sum()` as *Not Threatened*.

We obtained automated conservation assessments for `r aa %>% dplyr::select(flag_Category_CriteriaB) %>% filter(!is.na(flag_Category_CriteriaB)) %>% nrow()` species in the filtered dataset. Based on the filtered data, the automated conservation assessment evaluated `r sum(aa$flag_status == "possibly_threatened", na.rm = TRUE)` species as possibly threatened (`r round(sum(aa$flag_status == "possibly_threatened", na.rm=TRUE) / aa %>% filter(!is.na(flag_status)) %>% nrow() * 100, 1)`%, CR = `r aa %>% filter(flag_Category_CriteriaB == "CR") %>% nrow()`, EN = `r aa %>% filter(flag_Category_CriteriaB == "EN") %>% nrow()`, VU = `r aa %>% filter(flag_Category_CriteriaB == "VU") %>% nrow()`, see Appendix 3 for assessments of all species). The automated assessment based on the filtered dataset agreed with the IUCN assessment for identifying species as possibly threatened (CR, EN, VU) for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status)) %>% mutate(agreement = iucn_status == flag_status) %>% dplyr::select(agreement) %>% unlist() %>% sum() ` species (`r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status))  %>% mutate(agreement = iucn_status == flag_status) %>% dplyr::select(agreement) %>% unlist() %>% mean() %>%  round(3) * 100`%; Table 4). Filtering reduced the EOO by `r tab3 %>%  dplyr::select(flag_median_EOO_change) %>%  unlist() %>%  median() %>%  abs()`% and the AOO by `r tab3 %>%  dplyr::select(flag_median_AOO_change) %>%  unlist() %>%  median() %>% abs()`% on median per group. For the raw dataset the agreement with IUCN was higher at `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% mutate(agreement = iucn_status == raw_status) %>% dplyr::select(agreement) %>% unlist() %>% sum() ` species (`r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% mutate(agreement = iucn_status == raw_status) %>% dplyr::select(agreement) %>% unlist() %>% mean() %>%  round(3) * 100`%).

# Discussion {-}
<!-- add some subheadings to discussion -->
Automated flagging based on meta-data and automatic tests filtered on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  dplyr::select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% of the records per taxonomic group; `r min(tab2$total_unfit)`%-`r max(tab2$total_unfit)`% as "unfit" and `r min(tab2$total_error)`%-`r max(tab2$total_error)`% as "erroneous". The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (**Question 1**). The importance of different tests was similar across taxonomic groups, with particular exceptions for the tests on basis of record, collection year, capital coordinates, and urban areas (**Question 2**). The results for species richness were similar between the raw and filtered data with some improvements using the filters. We found little impact of filtering on the accuracy of the automated conservation assessments (**Question 3**).

@@

Overall, we consider the effect of the automated filters as positive since they identified the above-mentioned issues and increased the data precision and reduced computational burden (Table 3, Appendix 2). However, in some cases filters failed to remove major issues, often due to incomplete meta-data. For instance, for Diogenidae we found at least two records of an species known only from Eocene fossils (*Paguristes mexicanus*) which slipped the "basis of record" test because they were marked as "preserved specimen" rather than "fossil specimen". Furthermore, for Entomobryidae we found that for `r dat %>% filter(taxon == "Entomobryidae") %>%  filter(!is.na(species)) %>%  filter(taxonRank == "UNRANKED") %>%  nrow()` records the meta-data on taxonomic rank was "UNRANKED" despite all of them being identified to species level, leading to a high fraction of records removed by the "Identification level" filter. Additionally automated filters might be overly strict  or unsuitable for certain taxa. For instance, in Entomobryidae, `r dat %>% filter(taxon == "Entomobryidae") %>% filter( basisOfRecord == "MATERIAL_SAMPLE") %>% nrow()` samples were marked as material samples and therefore removed by our global filter retaining only specimen and observation data, which in this case was overly strict.

The filters we included in this study address a set of important but relatively easy to identify problems. In fact, the internal quality control of GBIF does flag some of the problems we tested for (i.e., zero coordinates, equal lat/lon) while others might be implemented in the near future (country centroids, https://data-blog.gbif.org/post/country-centroids/). While this internal quality is very helpful, we see a huge potential to overcome issues with data quality in a user-feedback system that allows users to provide expert assessments, i.e. a meta-annotation of records being challenged (and why). Such a system would not need to change the original data and could include multiple levels to account for differing opinions.

As next steps for automated filtering, tests for intrinsic consistency and support by external data (if available) can help to detect additional problematic records. For instance, testing if records' coordinates fall within the state or province of collection noted for a record (intrinsic) or testing if agree with external species distribution information, for example from www.iucn.org (vertebrates) or https://wcsp.science.kew.org/ (selected seed plant families; extrinsic) can further corroborate the accuracy of a record's geographic referencing. If such tests are included it is essential to account for the sampling year, in particular for older records, since the names of political entities may change and the ranges of species may shift. Furthermore, while in this study we focused on meta-data and geographic filtering, taxonomic cleaning---the resolution of synonymies and identification of accepted names---is another important part of data curation, but depends on taxon-specific taxonomic backbones and synonymy lists which are not readily available for many groups and often are contradictory within individual taxa.

@@

# Conclusions {-}
Our results suggest that between one quarter to half of the occurrence records obtained from GBIF might be unsuitable for downstream biodiversity analyses. While the majority of these records might not be erroneous *per se*, they are overly imprecise and thereby increase uncertainty of downstream results or add computational burden on big data analyses.
